On Optimality of Finite Element Integration

### 2015-12-16 TOMS_Paper/assembly.tex
Other small fixes

To illustrate the definition, we show in Figure~\ref{code:multi_loopnest} how sharing evolves as factorization and code motion are applied to a trivial multilinear loop nest. In the original loop nest (Figure~\ref{code:multi_loopnest_a}), spatial sharing is induced by $b_j$. Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}). Finally, generalized code motion (\cite{Luporini}) leads to optimality (Figure~\ref{code:multi_loopnest_c}).

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion. If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations. Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This require coordination with other transformations, as discussed in the following sections. For the moment -- and for ease of treatise -- we neglect this aspect.}, with optimality being achieved if stronger hypotheses hold. Relaxing the problem is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.

@@

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For simplicity, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}[label=(\roman*)]
\item Applying generalized code motion -- not needed in the example as we have one loop.
\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$. In general, however, the optimal option depends on both the expansion cost in (ii) and the loop size. One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}. We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.

Of notable importance, in the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations. Often, structured tensor operations characterize the $\sigma_{ei}$ sub-expressions too. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).

@@

The algorithm makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item $c_o^l$: the cost of an expression $o$ in a loop $L_l$ (i.e., operations in $o$ times $|L_l|$).
\item $n_o$: the number of occurrences of a sub-expression $o$ in the input expression.
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\end{itemize}

@@

%If there exist $P^i$, $P^j$, and $P^k$
%\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
%|\Lambda| + \sum_{s \in S_P} (n_{s} - 1)$. Note that:
%\begin{itemize}
%\item $|P||\Lambda|$ ($|S_P||\Lambda|$) represents the operation count in the innermost loop
%\item $n_o$ ($n_s$) captures the number of additions once $o$ ($s$) is factorized
%\end{itemize}
%We define the cost of $P$ as $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$.

@@

\begin{figure}[b]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
\end{CenteredBox}\caption{The loop nest produced by the algorithm for an input as in Figure~\ref{code:loopnest}.}\label{code:loopnest-opt}\end{figure}

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary. $\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis). By assumption, this case is neglected. The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one. We prevent this situation by construction. $\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}). Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of linearity, the operation count induced by a partition $P \in \mathbb{P}$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$. What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality. We do it by contradiction. Let us assume that there exists another ordering for $\mathbb{P}$ leading to $c_{\Lambda''_I} < c_{\Lambda'_I}$. In particular, this implies the existence of $P$, with $c_{P} > c_{P_{min}}$, such that if $P$ is scheduled prior to $P_{min}$, a lower operation count is achieved. We distinguish two cases.
\begin{itemize}
\item $|P| < |S_P|$. This is the simplest scenario, since generalized code motion can be applied to the multilinear operands in $P$ and no disruptive re-factorization takes place. We would then have $c_{\Lambda''_I} = c_{\Lambda'_I}$, which clearly contradicts the hypothesis.
\item $|P| \geq |S_P|$. The expression is re-factorized such that the multilinear symbols in $S_P$ can be collected. Assuming $|S_P| = n$, this process exposes $n$ sub-expressions amenable to code motion. It is possible that these $n$ sub-expressions include all of the multilinear operands (symbols) in $P_{min}$. Due to linearity, the innermost loop will suffer from an increase in operation count of $2n - 1$ ($n$ products, $n-1$ additions). Conversely, assuming $|S_{P_{min}}| = n - 1$, if we scheduled $P_{min}$ first, . 
\end{itemize}

>>>>

To illustrate the definition, we show in Figure~\ref{code:multi_loopnest} how sharing evolves as factorization and code motion are applied to a trivial multilinear loop nest. In the original loop nest (Figure~\ref{code:multi_loopnest_a}), spatial sharing is induced by $b_j$. Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}). Finally, generalized code motion \cite{Luporini} leads to optimality (Figure~\ref{code:multi_loopnest_c}).

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion. If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations. Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections. For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold. Relaxing the problem is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.

@@

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For example, suppose that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}[label=(\roman*)]
\item Applying generalized code motion -- not needed in the example as we have one loop.
\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$. In general, however, the optimal option depends on multiple factors: the loop size, the expansion cost, and the factorization gain. One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}. We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.

Of notable importance, from the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations. Often, structured tensor operations characterize the $\sigma_{ei}$ sub-expressions too. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).

@@

The algorithm makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\end{itemize}

@@

@@

\begin{figure}[t]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
\end{CenteredBox}\caption{The loop nest produced by the algorithm for an input as in Figure~\ref{code:loopnest}.}\label{code:loopnest-opt}\end{figure}

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary. $\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis). By assumption, this case is neglected. The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one. We prevent this situation by construction. $\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}). Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of linearity, the operation count induced by a partition $P \in \mathbb{P}$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$.

%What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality. We do it by contradiction. Let us assume that there exists another ordering for $\mathbb{P}$ leading to $c_{\Lambda''_I} < c_{\Lambda'_I}$. In particular, this implies the existence of $P$, with $c_{P} > c_{P_{min}}$, such that if $P$ is scheduled prior to $P_{min}$, a lower operation count is achieved. We distinguish two cases.
%\begin{itemize}
%\item $|P| < |S_P|$. This is the simplest scenario, since generalized code motion can be applied to the multilinear operands in $P$ and no disruptive re-factorization takes place. We would then have $c_{\Lambda''_I} = c_{\Lambda'_I}$, which clearly contradicts the hypothesis.
%\item $|P| \geq |S_P|$. The expression is re-factorized such that the multilinear symbols in $S_P$ can be collected. Assuming $|S_P| = n$, this process exposes $n$ sub-expressions amenable to code motion. It is possible that these $n$ sub-expressions include all of the multilinear operands (symbols) in $P_{min}$. Due to linearity, the innermost loop will suffer from an increase in operation count of $2n - 1$ ($n$ products, $n-1$ additions). Conversely, assuming $|S_{P_{min}}| = n - 1$, if we scheduled $P_{min}$ first, . 
%\end{itemize}
